---
title: "christina_EDA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(tidyverse)
```

## Introduction


```{r}
matches <- read.csv("data/international_matches.csv")
box <- read.csv("data/FIFAallMatchBoxData.csv")
fifa_rankings<-read.csv("data/fifa_ranking-2022-10-06.csv")
```


```{r}
matches <- matches %>% 
  mutate(year = substr(date,1,4))

box$year <- as.character(box$year)
  

final_dataset <- inner_join(matches, box, by = c( "year"= "year",
                                            "home_team" = "hname", 
                                            "away_team" = "aname", 
                                            "home_team_score" = "hgoals", 
                                            "away_team_score" = "agoals"))

```

```{r}
final_dataset%>%
  arrange(desc(date))
```


```{r}
ggplot(data = final_dataset) +
  geom_histogram(mapping = aes(x = home_team_score), binwidth = 0.5)
```

```{r}
ggplot(data = final_dataset) +
  geom_histogram(mapping = aes(x = away_team_score), binwidth = 0.5)
```
```{r}
ggplot(data = final_dataset) +
  geom_histogram(mapping = aes(x = away_team_score), binwidth = 0.5)
```



```{r}
ggplot(data = final_dataset, mapping = aes(x = away_team_score, y = away_team_fifa_rank)) + 
  geom_point()
```


Top 10 FIFA ranked teams 

```{r}
home_team_ranks <- final_dataset%>%
  select(home_team,date, home_team_fifa_rank)%>%
  rename(
    team = home_team, 
    fifa_rank = home_team_fifa_rank
  )

away_team_ranks <- final_dataset%>%
  select(away_team, date, away_team_fifa_rank)%>%
  rename(
    team = away_team, 
    fifa_rank = away_team_fifa_rank
  )

team_ranks <-rbind(home_team_ranks, away_team_ranks)
  
#  merge(x=home_team_ranks,y=away_team_ranks, 
  #           by=c("team","date","fifa_rank"))
top_team_ranks<- team_ranks%>%
    arrange((team),desc(date)) %>%
    group_by(team)%>%
    slice(1) 

#%>%
 #   arrange(fifa_rank)%>% 
 #   slice(1:10)
  

fifa_top_10<-top_team_ranks%>%
  arrange(fifa_rank)%>%head(10)

fifa_top_10
```


```{r}
fifa_rankings%>%
  filter(rank_date=="2022-10-06")%>%
  arrange(rank)%>%
  slice(1:10)
```


```{r}
final_dataset<- final_dataset%>%
  mutate(goal_differential = home_team_score-away_team_score,
         rank_differential =away_team_fifa_rank-home_team_fifa_rank)
```


```{r}
ggplot(data = final_dataset) +
  geom_histogram(mapping = aes(x = goal_differential), binwidth = 0.5)
```


Home team and away team rank differential vs home team rank differential

```{r}
ggplot(data = final_dataset, mapping = aes(x = goal_differential, y =rank_differential)) + 
  geom_point()
```
The greater the difference in rank (meaning the higher the home team is ranked against the away team), the higher the goal differential (the more the home team scores against the away team). 

Home team and away team rank differential vs home team result (win, draw, loss)

```{r}
ggplot(data = final_dataset, mapping = aes(x = rank_differential, y =home_team_result)) + 
   geom_boxplot()
```

Possesion vs home team result

```{r}
ggplot(data = final_dataset, mapping = aes(x = hPossesion, y =home_team_result)) + 
   geom_boxplot()
```



Best 10 offensive teams 

```{r}
home_team_offense <- final_dataset%>%
  select(home_team,date, home_team_mean_offense_score)%>%
  rename(
    team = home_team, 
    offense_score = home_team_mean_offense_score
  )

away_team_offense <- final_dataset%>%
  select(away_team, date, away_team_mean_offense_score)%>%
  rename(
    team = away_team, 
    offense_score = away_team_mean_offense_score
  )

team_offense <-rbind(home_team_offense, away_team_offense)
  
#  merge(x=home_team_ranks,y=away_team_ranks, 
  #           by=c("team","date","fifa_rank"))
top_team_offense<- team_offense%>%
    arrange((team),desc(date)) %>%
    group_by(team)%>%
    slice(1) 

top_10_offense <- top_team_offense%>%
  arrange(desc(offense_score))%>%head(10)

top_10_offense
```


Best 10 midfield teams 
```{r} 
home_team_midfield <- final_dataset%>%
  select(home_team,date, home_team_mean_midfield_score)%>%
  rename(
    team = home_team, 
    midfield_score = home_team_mean_midfield_score
  )

away_team_midfield <- final_dataset%>%
  select(away_team, date, away_team_mean_midfield_score)%>%
  rename(
    team = away_team, 
    midfield_score = away_team_mean_midfield_score
  )

team_midfield <-rbind(home_team_midfield, away_team_midfield)
  
top_team_midfield<- team_midfield%>%
    arrange((team),desc(date)) %>%
    group_by(team)%>%
    slice(1) 

top_10_midfield <- top_team_midfield%>%
  arrange(desc(midfield_score))%>%head(10)


top_10_midfield
```

Best 10 defensive teams

```{r}
home_team_defense <- final_dataset%>%
  select(home_team,date, home_team_mean_defense_score)%>%
  rename(
    team = home_team, 
    defense_score = home_team_mean_defense_score
  )

away_team_defense <- final_dataset%>%
  select(away_team, date, away_team_mean_defense_score)%>%
  rename(
    team = away_team, 
    defense_score = away_team_mean_defense_score
  )

team_defense <-rbind(home_team_defense, away_team_defense)
  
#  merge(x=home_team_ranks,y=away_team_ranks, 
  #           by=c("team","date","fifa_rank"))
top_team_defense<- team_defense%>%
    arrange((team),desc(date)) %>%
    group_by(team)%>%
    slice(1) 

top_10_defense <- top_team_defense%>%
  arrange(desc(defense_score))%>%head(10)

```


Top 10 teams with highest wining percentage at home and away


Correlatoin Matrix 

