---
title: "sean_EDA"
author: "Sean Li"
date: '2022-12-03'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
```

## Introduction

```{r}
matches <- read.csv("data/international_matches.csv")
box <- read.csv("data/FIFAallMatchBoxData.csv")
```

naming stuff
```{r}
matches <- matches %>% 
  mutate(year = substr(date,1,4))

box$year <- as.character(box$year)
```

```{r}
matches[matches == "Côte d'Ivoire"] <- "Ivory Coast"
matches[matches == "IR Iran"] <- "Iran"
matches[matches == "Korea Republic"] <- "South Korea"
matches[matches == "China PR"] <- "China"
matches[matches == "USA"] <- "United States"
```


```{r}
final_dataset <- inner_join(matches, box, by = c( "year"= "year",
                                            "home_team" = "hname", 
                                            "away_team" = "aname", 
                                            "home_team_score" = "hgoals", 
                                            "away_team_score" = "agoals"))
```



```{r}
wc_years = list(2006, 2002, 2010,2014,2018,2022)
final_dataset %>% 
  filter(year %in% wc_years)
```

```{r}
unique(box$year)
```


# getting last 5 for Iran, Qatar, Saudi Arabia, Morocco, Canada



```{r}
cm <- read.csv("data/cleaned_matches.csv")
library(readxl)
missing <- read_xlsx("data/last5.xlsx")
```


## Another attempt at modeling

###Multinomial Logistic Regression. Predicting Win/Loss/Draw. For moneyline.

```{r}
#train-test split
library(nnet)
#factor the result, draw is baseline
train_data_biv$home_team_result <- as.factor(train_data_biv$home_team_result) 

sample <- sample(c(TRUE, FALSE), nrow(train_data_biv), replace=TRUE, prob=c(0.7,0.3))
traindf  <- train_data_biv[sample, ]
testdf   <- train_data_biv[!sample, ]
```


```{r}
#actual model + summary
log_model <- multinom(home_team_result ~ fifa_rank_differential + offense_differential + goalkeeper_differential+
          neutral_location+  midfield_differential+ defense_differential+
       shots_on_target_differential+shots_differential+yellow_cards_differential+fouls_differential, data = traindf, model = TRUE)
summary(log_model)
```


Accuracy

```{r}
# Predicting the values for train dataset
traindf$result_pred <- predict(log_model, newdata = traindf, "class")

# Building classification table
train_tab <- table(traindf$home_team_result, traindf$result_pred)

# Calculating accuracy - sum of diagonal elements divided by total obs
print(paste0("training accuracy: " , round((sum(diag(train_tab))/sum(train_tab))*100,2)))
train_tab
```

```{r}
# Predicting the values for train dataset
testdf$result_pred <- predict(log_model, newdata = testdf, "class")

# Building classification table
test_tab <- table(testdf$home_team_result, testdf$result_pred)

# Calculating accuracy - sum of diagonal elements divided by total obs
print(paste0("test accuracy: " , round((sum(diag(test_tab))/sum(test_tab))*100,2)))
test_tab
```
predicted is on the top, actual is on the side


now testing on the actual world cup matches

```{r}


actualworldcup <- final_model2_data %>%  mutate(home_team_result = 
                                                  case_when(country1_score-country2_score>0 ~ "Win", 
                                                           country1_score-country2_score == 0 ~ "Draw",
                                                           country1_score-country2_score<0 ~ "Lose")) %>% select(home_team_result,fifa_rank_differential,offense_differential, goalkeeper_differential,neutral_location,midfield_differential,defense_differential,
       shots_on_target_differential,shots_differential,yellow_cards_differential,fouls_differential)
actualworldcup <- actualworldcup %>% mutate(home_team_result = factor(home_team_result)) %>% 
  mutate(home_team_result = fct_relevel(home_team_result, c("Draw","Lose","Win"))) %>% 
  mutate(neutral_location = case_when(neutral_location~0, TRUE~1)) #%>% 
  #arrange(home_team_result)


 

# Predicting the values for train dataset
actualworldcup$result_pred <- predict(log_model, newdata = actualworldcup, "class")

# Building classification table
test_tab <- table(actualworldcup$home_team_result, actualworldcup$result_pred)

# Calculating accuracy - sum of diagonal elements divided by total obs
print(paste0("test accuracy: " , round((sum(diag(test_tab))/sum(test_tab))*100,2)))
test_tab
```
### Poisson Model for # of goals for any team

```{r}
# Building train and test set for CV
set.seed(1)

n <- nrow(train_data_biv)


train_nums <- seq(1, 0.8* nrow(cups_only_nona), 1)

#all the X variables for test data
test_data = cups_only_nona[-train_nums, ] %>%
  select(rank_differential, offense_differential, defense_differential, midfield_differential,
         points_differential, neutral_location) # add cap_diff if using
#predict function needs a mat, makes categorical into dummy vars
mat = model.matrix(~., test_data)
# actual y - values
test_home <- cups_only_nona[-train_nums, "home_team_score"] %>%
  pull()
test_away <- cups_only_nona[-train_nums, "away_team_score"] %>% 
  pull()

train_data <- cups_only_nona[train_nums, ]


# Model equations 
home_x <- model.matrix(home_team_score ~ rank_differential + offense_differential + 
                         defense_differential + midfield_differential + 
         points_differential + neutral_location, 
                       data = train_data) # add cap_diff if using
away_x <- model.matrix(away_team_score ~  + rank_differential + offense_differential + 
                         defense_differential + midfield_differential + 
         points_differential + neutral_location, 
                       data = train_data) # add cap_diff if using

home_y <- train_data$home_team_score
away_y <- train_data$away_team_score


# Cross validating 
lambda_grid <- 10^seq(-2, 10, length.out = 5000)
home_model <- cv.glmnet(home_x, home_y, alpha = 1, 
                        lambda = lambda_grid, family = "poisson",
                        nfolds = 5)
away_model <- cv.glmnet(away_x, away_y, alpha = 1, lambda = lambda_grid, family = "poisson", 
                        nfolds = 5)

home_lambda <- home_model$lambda.min
away_lambda <- away_model$lambda.min


# Fitting final models

final_home <- glmnet(home_x, home_y, alpha = 1, lambda = home_lambda, family = "poisson")
final_away <- glmnet(away_x, away_y, alpha = 1, lambda = away_lambda, family = "poisson")

final_home$beta
final_away$beta

# Generating preds
home_preds <- predict(home_model, newx = mat, type = "response")
away_preds <- predict(away_model, newx = mat, type = "response")
```


# Model Diagnostics: 

In order to use our model, we successfully checked the model diagnostics as follows:
1. Make sure that the counts are poisson distributed. The mean = variance for goals scored. 
2. There are linear relationships between log(response) and changes in predictor variables. Details are in the appendix. 

#odds

```{r}
results_and_odds <-read_xlsx("data/group_pred_with_odds.xlsx")

#convert odds to numerical probability
results_and_odds <- results_and_odds %>% 
  mutate(winProb = case_when(winOdds>0 ~ (100/(winOdds+100)), TRUE ~ (-winOdds/(-winOdds+100))),
         drawProb = case_when(drawOdds>0 ~ (100/(drawOdds+100)), TRUE ~ (-drawOdds/(-drawOdds+100))),
         loseProb = case_when(loseOdds>0 ~ (100/(loseOdds+100)), TRUE ~ (-loseOdds/(-loseOdds+100)))
         )

```


# Conclusion and Connection to Sports Betting

• Conclusions & findings fully supported by statistical analysis and data
• Conclusions & findings fully address the study objectives in a thoughtful & nuanced way
• Statistical analyses are translated into meaningful and understandable findings for the target audience (e.g., managers, engineers)
• Any limitations of approach are discussed, or conclusions reached are warranted by the modeling approach


In conclusion, our models tackle two of the most important bets in soccer, the 3-way moneyline (win, loss, draw) as well as the goal differential. For the group stage in the 2022 World Cup, we came up with relative probabilities for match outcome based on modeling the number of home goals and away goals on a bivariate poisson model. We then decided to see if our model could make money based on the pregame odds for goal differential as well as moneyline. 

For moneyline bets, our model would have made $X over the group stage. We first obtain the odds for each group stage game. Then we converted the odds to numerical probability. Then we compared the probabilities to those games in which there was a clear difference in expected outcome, and bet on that difference. 

Sports bettors should heed caution though when using our model, as the draw accuracy is relatively low compared to win or loss. For X/48 matches that ended as draws, it only predicted 0/48 matches to end as a draw. Therefore moneyline betters should take caution on the odds we are predicting. 

A few limitations do exist though. Sports are inherently hard to predict, which is why any model we built did not have accuracy above 60%. If building high performing models was easy, then sports books would go out of business. Furthermore, it is difficult to take into account the interaction between two specific teams as they prepare customized game plans that differ based upon opponent. These were not taken into account in our predictors for the models were built. Future work may include building models to target other bets, such as over/under on total goals scored or various team/player prop bets. 