---
title: "World Cup 2022"
author: "Sean Li, Brian Janger, Aaditya Warrier, Christina Yoh"
date: '2022-12-03'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Methodology: 
Using historical data and group stage matches individual match up to the final 16 (because we know these matchups will occur) 
Based on this data, we will calculate win probabilities, loss probabilities, and expected score 
Based on individual matches in the group stage up to round of 16, we will calculate who wins top of their group/2nd etc. 

Model: 
Use a probabilistic model, because sports betters want to know probabilities 
Poisson GLM, can use lasso
Monte Carlo 

We are predicting the goal-line (MUST BE ABSOLUTE VALUE CAUSE POISSON WILL NOT WORK OTHERWISE) for each match upto round of 16 using a Poisson GLM that takes home stats and away stats into account, as well as individual team rosters 

Will need to make two diff models for home and away because of negative value restriction

## FINAL WRITEUP

```{r}
library(tidyverse)
library(DescTools)
library(glmnet)
library(extraDistr)
data("d.countries")
source("biv-pois-functions/lm.bp.R")
source("biv-pois-functions/bivpois.table.R")
source("biv-pois-functions/newnamesbeta.R")
source("biv-pois-functions/pbivpois.R")
source("biv-pois-functions/simple.bp.R")
source("biv-pois-functions/splitbeta.R")
int_matches <- read_csv("data/international_matches.csv")
qatar_teams <- read_delim("data/Qatar2022-teams.csv", delim = ";")
matches_sched <- read_delim("data/matchs-schedule.csv", delim = ";")
final_roster_data <- read_csv("data/final_roster_data.csv")
```


```{r}
# should we use fifa only matches? we only have roster for these, still about 5000 matches including # qualifying mmatches
# fifa_only <- int_matches %>%
#   filter(str_detect(tournament, "FIFA"))

# DECENT UNIVARIATE MODELS???


cups_only <- int_matches %>%
  filter(!str_detect(tournament, "Friendly|qualification"), format(date, "%Y") >= 2002) %>%
  select(-home_team_goalkeeper_score, away_team_goalkeeper_score) %>%
  mutate(rank_differential = home_team_score - away_team_score,
         offense_differential = home_team_mean_offense_score - away_team_mean_offense_score,
         defense_differential = home_team_mean_defense_score - away_team_mean_defense_score,
         midfield_differential = home_team_mean_midfield_score - away_team_mean_midfield_score,
         points_differential = home_team_total_fifa_points - away_team_total_fifa_points)

# cross validating

set.seed(2)
cups_only_nona <- na.omit(cups_only)
n <- nrow(cups_only_nona)
train_nums <- sample(1:n, size = 0.9*n)

test_data = cups_only_nona[-train_nums, ] %>%
  select(rank_differential, offense_differential, defense_differential, midfield_differential,
         points_differential, neutral_location, shoot_out)

test_home <- cups_only_nona[-train_nums, "home_team_score"] %>%
  pull()
test_away <- cups_only_nona[-train_nums, "away_team_score"] %>% 
  pull()

train_data <- cups_only_nona[train_nums, ]

home_x <- model.matrix(home_team_score ~ rank_differential + offense_differential + 
                         defense_differential + midfield_differential + 
         points_differential + neutral_location + shoot_out, 
                       data = train_data)
away_x <- model.matrix(away_team_score ~  + rank_differential + offense_differential + 
                         defense_differential + midfield_differential + 
         points_differential + neutral_location + shoot_out, 
                       data = train_data)

home_y <- train_data$home_team_score
away_y <- train_data$away_team_score


lambda_grid <- 10^seq(-2, 10, length.out = 5000)
home_model <- cv.glmnet(home_x, home_y, alpha = 1, lambda = lambda_grid, family = "poisson",
                        nfolds = 5)
away_model <- cv.glmnet(away_x, away_y, alpha = 1, lambda = lambda_grid, family = "poisson", 
                        nfolds = 5)

home_lambda <- home_model$lambda.min
away_lambda <- away_model$lambda.min

final_home <- glmnet(home_x, home_y, alpha = 1, lambda = home_lambda, family = "poisson")
final_away <- glmnet(away_x, away_y, alpha = 1, lambda = away_lambda, family = "poisson")

final_home$beta

final_away$beta

mat = model.matrix(~., test_data)


home_preds <- predict(home_model, newx = mat, type = "response")
away_preds <- predict(away_model, newx = mat, type = "response")


```


```{r checking}
# SANITY CHECKS 
sqrt(mean((test_home - home_preds)^2))
sqrt(mean((test_away - away_preds)^2))

abs(median(test_home - home_preds))
abs(median(test_away - away_preds))


goalline_preds <- home_preds - away_preds

outcomes <- na.omit(cups_only)[-train_nums, "home_team_result"] 

outcomes %>%
  count(home_team_result)

sum(goalline_preds  <0)
```


```{r FINAL BIVPOIS MODEL}
biv_cups_only_nona <- cups_only_nona %>%
  mutate(shoot_out = ifelse(shoot_out == "Yes", 1, 0),
         neutral_location = ifelse(neutral_location == TRUE, 1, 0))

# need to test accuracy still of bivpoiss with train-test - looks p great though
set.seed(2)
n <- nrow(biv_cups_only_nona)
train_nums <- sample(1:n, size = 0.9*n)


fin_mod <- lm.bp(l1 = home_team_score~ rank_differential + points_differential, 
                 l2 = away_team_score~ rank_differential + offense_differential + 
                   points_differential + neutral_location + shoot_out,
                 l3 = ~rank_differential, zeroL3 = TRUE, data = biv_cups_only_nona)
```

```{r monte carlo}

# Monte Carlo Simulation 

# use this to generate data - https://stats.stackexchange.com/questions/27443/generate-data-samples-from-poisson-regression
```

